---
title: "R Notebook"
output: html_notebook
---

# Exercise Sheet 1

## Ex1

A call centre receives calls according to a Poisson process. The expected number of calls it receives in a day is 1523. On any given day, the call centre has enough staff to cope with a maximum of 1600 calls. What is the probability that on any given day the call centre receives more calls than it can cope with.

```{r}
lambda = 1523
1 - ppois(1600, lambda)
```

## Ex2

Write code that simulates a homogeneous temporal Poisson process on $(0, T]$ with intensity λ. Have a go at writing it as a function homogeneouspp(lambda,T), such that it can take any value of λ and T that you wish to give it.

```{r}
homogeneouspp = function(lambda, T) {
  N = rpois(1, lambda * T)
  E = runif(N, min = 0, max = T)
  E = sort(E)
  return(E)
}
```

Simulate 100 realisations of a Poisson process on (0, 10] with intensity λ = 10.

1.  Verify that the sample mean number of events in the interval (3, 5] is consistent with the expected number of events in (3, 5].

```{r}
sum = 0
all_simu = c()
for (i in 1:100) {
  N = homogeneouspp(10, 10)
  len = length(which(N > 3 & N <= 5))
  sum = sum + len
}
sum/100
```

```{r}
all_simu
```

```{r}
N = c()
for (i in 1:100) {
  sim = homogeneouspp(10, 10)
  len = length(which(sim > 2 & sim <= 8))
  N[i] = len
}

hist(N)
plot(dpois(1:100, 60))
```

3.  
