---
title: "R Notebook for Problem Class 1"
output: html_notebook
---

# 4.1.1 Visualisation

## Read in the data

```{r}
MyData <- read.csv("EUR-GBP-Rates.csv", header = TRUE, sep = ",", dec = ".",fileEncoding="UTF-8-BOM")
```

## View the data

```{r}
View(MyData)
```

## Determine the number of observations in MyData

```{r}
nrow(MyData)
```

str function will display the internal structure of an R object

```{r}
str(MyData)
```

## print all the values of the GBPtoEUR

```{r}
MyData$GBPtoEUR
MyData[, 3]
```

## Some plotting

```{r}
plot(MyData$GBPtoEUR, type = 'l', col = 'blue', main = "EUR/GBP exchange rate", xlab = "Time", ylab = "GBP per EUR", cex.main = 1.25, cex.lab = 1.5, cex.axis = 2)
```

-   **type:** Change the line type

-   **col:** Change the color of the graph

-   **main:** title of the plot

-   **xlab:** label of the x-axis

-   **ylab:** label of the y-axis

-   **cex.main:** size of the plot title

-   **cex.lab:** font size of the labels of the axes

-   **cex.axis:** Size of the tick labels

### compare your initial plot and your modified plot side by side using the function par

```{r}
par(mfrow=c(1,2))
MyDates = as.Date(MyData$Date, "%d/%m/%Y")
plot(MyData$GBPtoEUR)
plot(MyDates, MyData$GBPtoEUR, type="l",col="blue", main="EUR/GBP
exchange rate", xlab="Time", ylab="GBP per one EUR", cex.main=1.25, cex.lab=1.5, cex.axis=2)
par(mfrow=c(1,1))
```

# 4.1.2 Sample moments and empirical distribution

## Sample mean

$$
\mathrm{E}(X)=\left\{\begin{array}{ll}\sum_{k} k \mathrm{P}(X=k) & \text { if } X \text { is discrete } \\\int_{-\infty}^{\infty} x f_{X}(x) d x & \text { if } X \text { is continuous. }\end{array}\right.
$$

Try some R code!

```{r}
mean(MyData$GBPtoEUR)
```

## Sample variance

The variance of a random variable $X$ with mean $\mu=\mathrm{E}(X)$ is defined as

$$
\text{Var}(X) = \operatorname{E}[(X-\mu)^2]
$$

```{r}
acf(MyData$GBPtoEUR, type = "covariance", plot = FALSE)
```

## Empirical distribution

```{r}
plot.ecdf(MyData$GBPtoEUR)
```

# 4.1.3 Bonus material: Histograms and (empirical) quantiles

## Histograms and empirical density

```{r}
hist(MyData$GBPtoEUR)
```

```{r, fig.height=6, fig.width=9}
par(mfrow=c(3,3))
hist(MyData$GBPtoEUR,breaks=4)
hist(MyData$GBPtoEUR,breaks=8)
hist(MyData$GBPtoEUR,breaks=12)
hist(MyData$GBPtoEUR,breaks=16)
hist(MyData$GBPtoEUR,breaks=20)
hist(MyData$GBPtoEUR,breaks=24)
hist(MyData$GBPtoEUR,breaks=28)
hist(MyData$GBPtoEUR,breaks=32)
hist(MyData$GBPtoEUR,breaks=36)
par(mfrow=c(1,1))
```

## Quantiles

```{r}
summary(MyData$GBPtoEUR)
```
